Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2022-06-03 manuscript.Rmd
cite concerns with preregistration

One reason why researchers are hesitant to preregister is that they feel restricted by the boundaries specified in their preregistration.
Even researchers who have been persuaded of its advantages sometimes apply strategies to skirt these self-imposed restrictions.
One strategy is to write a loose preregistration, to begin with; another is to deviate from the preregistration afterwards.
To complicate matters further, both strategies may be used with compelling scientific reasoning or with the self-serving intent of generating desirable results no matter the nature of the phenomenon under study.
To summarize, the conundrum is that the current conceptualization of preregistration deems a study "properly" preregistered even if it contains a recipe for p-hacking before data collection but "improper" if the researcher did technically deviate afterwards.
This, of course, goes against researchers' common intuition about what is considered exploratory and confirmatory.
We take these issues as signals of a misunderstanding about how preregistration contributes to scientific progress.
In particular, we want to challenge the common view that preregistration somehow cleanly separates confirmatory from exploratory.

@@

To arrive at this interpretation, we rely on three arguments.
The first is that theoretical risk is vital for judging theories.
This argument comes natural to a (Neo‑)Popperian or falsificationist, but  is not self-evident for a conformationist. (WHY; A BIT MORE MORE ELABORATION)
The second argument is that the theoretical risk for a given study is generally uncertain.
The third and last assumption is that this uncertainty is reduced by applying preregistration.

@@

Let us start by defining what we mean by expected epistemic value. (METADISCOURSE UNNECESSARY)
If researchers plan to conduct a study, they usually hope it will change their assessment of some theory's verisimilitude (truthlikeness).
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study concerning the verisimilitude of a specific theory is what we call epistemic value.
While researchers can not know what exactly they will learn from a study, they can form an expectation that helps them decide which study to conduct.
This expectation is what we term expected epistemic value.
We assume three things about this estimation process and how it relates to choosing a study to conduct.

@@

The first assumption leads to our adoption of a Bayesian framework. (WHY)
Our rationale is as follows.
Researchers that decide to conduct a study are akin to choosing a study to bet on.
They have to "place the bet" by conducting the study, therefore, invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study as a betting problem allows us to apply a "Dutch Book" argument [@christensenCleverBookiesCoherent1991].
This argument states that any better must follow the axioms of probability to avoid being "irrational", i.e., accepting bets that lead to sure losses.
Fully developing a Dutch book argument for this problem requires careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modelling what liberties researchers have in choosing a study.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are sufficient to derive the Bayes formula, on which we will heavily rely for our further arguments.
The argument is not sufficient, however, to warrant conceptualizing the kind of epistemic value we reason about in terms of probability; that remains a leap of faith.
Please note that our decision to adopt this aspect of the Bayesian philosophy of science does not imply anything about the statistical methods researchers use.
In fact, this conceptualization is purposefully reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.

@@

Let us make some simplifying assumptions and define notation. (METADISCOURSE UNNECESSARY)
We restrict ourselves to evidence of a binary nature (either exists or does not) since continuous evidence would lead to some quite involved derivations.
We denote the probability of a hypothesis before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with help from the Bayes formula:

@@

$P(H|E)$ is of great relevance since $P(H|E)$ is often used directly or indirectly as a measure of corroboration of a hypothesis.
In the tradition of Carnap (xxx), in its direct use, it is called corroboration as firmness; in its relation to $P(H)$, it is called increase in firmness. (SENTENCE STRUCTURE)
We refrain from discussing specific measures of corroboration since no measure shows universally better properties than others.
However, it is generally expected that any measure of corroboration increases monotonically with an increase in $P(H|E)$.

In short, we want to increase $P(H|E)$.
Increases in $P(H|E)$ are associated with increased epistemic value, of which we want to maximize the expectation.
So how can we increase $P(H|E)$?
The Bayes formula gives us three options to investigate, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher $P(H)$ leads to higher $P(H|E)$. However, the a priori probability of a hypothesis is nothing our study design can change.
The second option is similar commonsensical; that is, an increase in $P(E|H)$ leads to higher $P(H|E)$.
Consequently, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
While this truism is strongly related to study design, it is unrelated to preregistration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, increasing it will decrease $P(H|E)$: The more unlikely it is to observe evidence, the more it increases the probability of the hypothesis if we do observe it. (REPHRASE SENTENCE)
In other words, high risk, high reward. (REPHRASE SENTENCE)

If we equate riskiness with a low probability of obtaining evidence, the Bayesian rationale perfectly aligns with the observation that risky predictions lead to persuasive evidence.
This tension between high risk leading to high reward is central to our consideration of preregistration.
A high risk, high reward strategy is bound to result in many losses that are eventually absorbed by the high gains.
Sustaining many "failed" studies is not exactly aligned with the incentive structure under which many if not most researchers operate.
Consequently, researchers have an incentive to appear to take more risks than they actually do, which misleads their peers to give their claims more credence than they deserve.
It is at this juncture that the practice and mispractice of preregistration comes into play.
They try to ensure a proper judgment of the riskiness of a study by making tempering with the appearance of $P(E)$ difficult. (REPHRASE; THIS SENTENCE IS NON-SENSICAL)

@@

Let us note some interesting properties of theoretical risk $P(¬H|¬E)$. (WHY? // THIS IS NOT WELL INTRODUCED)
First, increasing theoretical risk leads to higher $P(H|E)$ (our objective).
Second, if the theoretical risk is smaller than $P(E|H)$ it follows that $P(H|E)$ must decrease when observing the evidence.
Third, if the theoretical risk equals zero, then $P(H|E)$ can at best be equal to $P(H)$ but only if $P(H|E)$ is one.
In other words, observing a sure fact does not lend credence to a hypothesis.

@@

First, the theory itself limits theoretical risk:
Some theories simply do not make risky predictions, and preregistration will not change that.
It would show a great deal of arrogance to advise researchers to stop investigating such theories. (META-DISCOURSE; UNNECESSARY LINE OF ARGUMENTATION?)
Consider the case of a researcher contemplating the relation between two sets of variables.
Suppose each set by itself is well studied, and strong theories tell the researcher how the variables within the set relate.
However, our imaginary researcher considers the relation between these two sets.
For lack of a better theory, they assume that some relation between any variables of the two sets exists.
This is not a risky prediction to make in psychology, even without statistical issues like alpha inflation [@orbenCrudReDefined2020].
However, we would consider it a success if the researcher would use the evidence to develop a more precise (and therefore risky) theory, e.g. by specifying which variables from one set relate to which variables from the other set, to what extent, in which direction, with which functional shape etc.
We will later show that preregistration increases the expected epistemic value of the further specified theory, though it remains low till substantiated by testing it again.
The point, however, is that we want to show that preregistration increases the expected epistemic value without regard to the theory being tested.
It is outside the scope of our agenda to changing or place demands on what theories should be tested through empirical research.

@@

This objective also raises the problem of changing the plan after data collection started. (WHY? THIS IS NOT WELL INTRODUCED)
One could argue that allowing deviation from the preregistration would allow researchers the same freedoms as having no preregistration at all.
Hypothethically at least, researchers might try any number of possible deviations and pick the one that is most favorable to their theory.
Hence, allowing deviation would void any otherwise guaranteed lower limits to the theoretical risk.
It follows that allowing deviations is wholly incompatible with the goal of maximizing theoretical risk.

@@

We established that higher theoretical risk leads to more credible evidence.
In other words, we have reconstructed the interpretation that preregistrations supposedly work by restricting the researchers, which in turn increases the theoretical risk (or equivalently limit type I error rate) and thereby creates more persuasive evidence.
Nevertheless, there are trade-offs for increasing theoretical risk.
However, employing a mathematical framework allows us to navigate the trade-offs more effectively and move towards a second, more favourable interpretation.
To that end, we incorporate uncertainty into our framework. (WELL DONE)

@@

Preregistration is often discussed in relation to statistical methods, so let us note some relations between $P(E|¬H)$ and statistical methods. (UNNECESSARY META-DISCOURSE)
If you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, then the nill-hypothesis is $¬H$, and therefore $P(E|¬H)$ is equivalent to the type I error rate.
Researchers who choose a smaller type I error rate can hence be more sure of their results, if significant.
However, the research hypothesis is seldomly equal to the statistical null hypothesis.
We argue that $P(E|¬H)$ also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.

@@

The point is that the contribution of statistical methods to $P(E|¬H)$ can be formally assessed. (META-DISCOURSE AVOID)
For many statistical models, their $P(E|¬H)$ can be analytically computed under some assumptions.
For those models or assumptions where this is impossible, one can employ Monte Carlo simulation to estimate the contribution to $P(E|¬H)$.
The precision with which researchers can discuss statistical contributions to $P(E|¬H)$ has lured the community concerned with research methods into ignoring other factors that are much more uncertain.
We can not hope to resolve this uncertainty; but we have to be aware of its implications. These are relayed in the following.

@@

Neither absolute certainty nor uncertainty are realistic scenarios but represent the boundary conditions that all realistic conditions fall .
Depending on the distribution that expresses our state of knowledge about the theoretical risk a study took on, our expectation for what to gain varies considerably.
Uncertainty about theoretical risk is expressed through the variance (or rather entropy) of the distribution.
Generally, we expect that increases in uncertainty (expressed as more entropic distributions) lead to a decreased expected epistemic value.

@@

Communicating clearly what the authors did to arrive at the evidence is crucial for judging the theoretical risk they took.
Preregistrations are ideal to communicate just that, because any description after the fact is suspect to be incomplete.
The authors could have decided to exclude a number of analytic strategies they tried out etc.
That is not to say that any study that was not preregistered was subjected to practices of p hacking.
The point is, we can not exclude this and a myriad of other possibilities and, hence, are left uncertain.
If the researcher do describe what they intend to do beforehand, and then report they did exactly that, the readers can be certain, that they have received a complete account of the situation.
They still might be uncertain about the actual theoretical risk the authors took but much less so. (REITERATE)
Remaining sources of uncertainty might be unfamiliarity with statistical methods or experimental paradigms, the probability of plain error in analyses etc.
In any case a well written preregistration should aim to reduce the uncertainty about the theoretical risk and hence increase the persuasiveness of evidence.

@@

The advantage of this perspective that puts uncertainty about theorethical risk front and center is twofold.
First, aiming to reduce uncertainty is beneficial across a wide range of potential theoretical risks.
That is, researchers benefit from preregistrating their study, regardless whether or not the study is "exploratory".
If researchers conduct a more exploratory study they can clearly communicate how exploratory they aim to be and their results can be judged accordingly.
Second, if researchers realize after the fact, that it would be unwise to follow their preregistration in certain details, they may change it.
Allowing to deviate from the preregistration is controversial (XXX) because the authors might sift through all the possibilities and than select the most beneficial for their theory.
This argument, in its fullest consequence, might result in the same uncertainty we argued is appropriate for not preregistered studies.
However, if the authors offer a convincing argument for their deviation, their readers might be compelled to increase their uncertainty only a limited amount. (RESTATE)

>>>>

One reason why researchers are hesitant to preregister is that they feel restricted by the boundaries specified in their preregistration [@goldin-meadowWhyPreregistrationMakes2016].
Even researchers who have been persuaded of its advantages sometimes apply strategies to skirt these self-imposed restrictions.
One strategy is to write a loose preregistration, to begin with; another is to deviate from the preregistration afterwards.
To complicate matters further, both strategies may be used with compelling scientific reasoning or with the self-serving intent of generating desirable results no matter the nature of the phenomenon under study.
To summarize, the conundrum is that the current conceptualization of preregistration deems a study "properly" preregistered even if it contains a recipe for p-hacking before data collection but "improper" if the researcher did technically deviate afterwards.
This, of course, goes against most researchers' intuition about what is considered exploratory and confirmatory.
We take these issues as signals of a misunderstanding about how preregistration contributes to scientific progress.
In particular, we want to challenge the common view that preregistration somehow cleanly separates confirmatory from exploratory.

@@

To arrive at this interpretation, we rely on three arguments.
The first is that theoretical risk is vital for judging theories.
This argument comes natural to a (Neo‑)Popperian or falsificationist, but  is not self-evident for a conformationist.
The second argument is that the theoretical risk for a given study is generally uncertain.
The third and last assumption is that this uncertainty is reduced by applying preregistration.

@@

Let us start by defining what we mean by expected epistemic value.
If researchers plan to conduct a study, they usually hope it will change their assessment of some theory's verisimilitude (truthlikeness).
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study concerning the verisimilitude of a specific theory is what we call epistemic value.
While researchers can not know what exactly they will learn from a study, they can form an expectation that helps them decide which study to conduct.
This expectation is what we term expected epistemic value.
We assume three things about this estimation process and how it relates to choosing a study to conduct.

@@

The first assumption leads to our adoption of a Bayesian framework.
Our rationale is as follows.
Researchers that decide to conduct a study are akin to choosing a study to bet on.
They have to "place the bet" by conducting the study, therefore, invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study as a betting problem allows us to apply a "Dutch Book" argument [@christensenCleverBookiesCoherent1991].
This argument states that any better must follow the axioms of probability to avoid being "irrational", i.e., accepting bets that lead to sure losses.
Fully developing a Dutch book argument for this problem requires careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modelling what liberties researchers have in choosing a study.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are sufficient to derive the Bayes formula, on which we will heavily rely for our further arguments.
The argument is not sufficient, however, to warrant conceptualizing the kind of epistemic value we reason about in terms of probability; that remains a leap of faith.
Please note that our decision to adopt this aspect of the Bayesian philosophy of science does not imply anything about the statistical methods researchers use.
In fact, this conceptualization is purposefully reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.

@@

Let us make some simplifying assumptions and define notation.
We restrict ourselves to evidence of a binary nature (either exists or does not) since continuous evidence would lead to some quite involved derivations.
We denote the probability of a hypothesis before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with help from the Bayes formula:

@@

$P(H|E)$ is of great relevance since $P(H|E)$ is often used directly or indirectly as a measure of corroboration of a hypothesis.
In the tradition of Carnap (XXX), in its direct use, it is called corroboration as firmness; in its relation to $P(H)$, it is called increase in firmness.
We refrain from discussing specific measures of corroboration since no measure shows universally better properties than others.
However, it is generally expected that any measure of corroboration increases monotonically with an increase in $P(H|E)$.

In short, we want to increase $P(H|E)$.
Increases in $P(H|E)$ are associated with increased epistemic value, of which we want to maximize the expectation.
So how can we increase $P(H|E)$?
The Bayes formula gives us three options to investigate, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher $P(H)$ leads to higher $P(H|E)$.
However, the a priori probability of a hypothesis is nothing our study design can change.
The second option is similar commonsensical; that is, an increase in $P(E|H)$ leads to higher $P(H|E)$.
Consequently, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
While this truism is strongly related to study design, it is unrelated to preregistration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, increasing it will decrease $P(H|E)$: The more unlikely it is to observe evidence, the more it increases the probability of the hypothesis if we do observe it.
In other words, high risk, high reward.

If we equate riskiness with a low probability of obtaining evidence, the Bayesian rationale perfectly aligns with the observation that risky predictions lead to persuasive evidence.
This tension between high risk leading to high reward is central to our consideration of preregistration.
A high risk, high reward strategy is bound to result in many losses that are eventually absorbed by the high gains.
Sustaining many "failed" studies is not exactly aligned with the incentive structure under which many if not most researchers operate.
Consequently, researchers have an incentive to appear to take more risks than they actually do, which misleads their readers to give their claims more credence than they deserve.
It is at this juncture that the practice and mispractice of preregistration comes into play.
They try to ensure a proper judgment of the riskiness of a study by making tempering with the appearance of $P(E)$ difficult.

@@

Let us note some interesting properties of theoretical risk $P(¬H|¬E)$.
First, increasing theoretical risk leads to higher $P(H|E)$ (our objective).
Second, if the theoretical risk is smaller than $P(E|H)$ it follows that $P(H|E)$ must decrease when observing the evidence.
Third, if the theoretical risk equals zero, then $P(H|E)$ can at best be equal to $P(H)$ but only if $P(H|E)$ is one.
In other words, observing a sure fact does not lend credence to a hypothesis.

@@

First, the theory itself limits theoretical risk:
Some theories simply do not make risky predictions, and preregistration will not change that.
It would show a great deal of arrogance to advise researchers to stop investigating such theories.
Consider the case of a researcher contemplating the relation between two sets of variables.
Suppose each set by itself is well studied, and strong theories tell the researcher how the variables within the set relate.
However, our imaginary researcher considers the relation between these two sets.
For lack of a better theory, they assume that some relation between any variables of the two sets exists.
This is not a risky prediction to make in psychology, even without statistical issues like alpha inflation [@orbenCrudReDefined2020].
However, we would consider it a success if the researcher would use the evidence to develop a more precise (and therefore risky) theory, e.g. by specifying which variables from one set relate to which variables from the other set, to what extent, in which direction, with which functional shape etc.
We will later show that preregistration increases the expected epistemic value of the further specified theory, though it remains low till substantiated by testing it again.
The point, however, is that we want to show that preregistration increases the expected epistemic value without regard to the theory being tested.
It is outside the scope of our agenda to changing or place demands on what theories should be tested through empirical research.

@@

This objective also raises the problem of changing the plan after data collection started.
One could argue that allowing deviation from the preregistration would allow researchers the same freedoms as having no preregistration at all.
Hypothethically at least, researchers might try any number of possible deviations and pick the one that is most favorable to their theory.
Hence, allowing deviation would void any otherwise guaranteed lower limits to the theoretical risk.
It follows that allowing deviations is wholly incompatible with the goal of maximizing theoretical risk.

@@

We established that higher theoretical risk leads to more credible evidence.
In other words, we have reconstructed the interpretation that preregistrations supposedly work by restricting the researchers, which in turn increases the theoretical risk (or equivalently limit type I error rate) and thereby creates more persuasive evidence.
Nevertheless, there are trade-offs for increasing theoretical risk.
However, employing a mathematical framework allows us to navigate the trade-offs more effectively and move towards a second, more favourable interpretation.
To that end, we incorporate uncertainty into our framework.

@@

Preregistration is often discussed in relation to statistical methods, so let us note some relations between $P(E|¬H)$ and statistical methods.
If you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, then the nill-hypothesis is $¬H$, and therefore $P(E|¬H)$ is equivalent to the type I error rate.
Researchers who choose a smaller type I error rate can hence be more sure of their results, if significant.
However, the research hypothesis is seldomly equal to the statistical null hypothesis.
We argue that $P(E|¬H)$ also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.

@@

The point is that the contribution of statistical methods to $P(E|¬H)$ can be formally assessed.
For many statistical models, their $P(E|¬H)$ can be analytically computed under some assumptions.
For those models or assumptions where this is impossible, one can employ Monte Carlo simulation to estimate the contribution to $P(E|¬H)$.
The precision with which researchers can discuss statistical contributions to $P(E|¬H)$ has lured the community concerned with research methods into ignoring other factors that are much more uncertain.
We can not hope to resolve this uncertainty; but we have to be aware of its implications.
These are relayed in the following.

@@

Neither absolute certainty nor uncertainty are realistic scenarios but represent the boundary conditions that all realistic conditions fall in.
Depending on the distribution that expresses our state of knowledge about the theoretical risk a study took on, our expectation for what to gain varies considerably.
Uncertainty about theoretical risk is expressed through the variance (or rather entropy) of the distribution.
Generally, we expect that increases in uncertainty (expressed as more entropic distributions) lead to a decreased expected epistemic value.

@@

Communicating clearly what the authors did to arrive at the evidence is crucial for judging the theoretical risk they took.
Preregistrations are ideal to communicate just that, because any description after the fact is suspect to be incomplete.
The authors could have decided to exclude a number of analytic strategies they tried out etc.
That is not to say that any study that was not preregistered was subjected to practices of p hacking.
The point is, we can not exclude this and a myriad of other possibilities and, hence, are left uncertain.
If the researcher do describe what they intend to do beforehand, and then report they did exactly that, the readers can be certain, that they have received a complete account of the situation.
They still might be uncertain about the actual theoretical risk the authors took but much less so.
Remaining sources of uncertainty might be unfamiliarity with statistical methods or experimental paradigms, the probability of plain error in analyses etc.
In any case a well written preregistration should aim to reduce the uncertainty about the theoretical risk and hence increase the persuasiveness of evidence.

@@

The advantage of this perspective that puts uncertainty about theorethical risk front and center is twofold.
First, aiming to reduce uncertainty is beneficial across a wide range of potential theoretical risks.
That is, researchers benefit from preregistrating their study, regardless whether or not the study is "exploratory".
If researchers conduct a more exploratory study they can clearly communicate how exploratory they aim to be and their results can be judged accordingly.
Second, if researchers realize after the fact, that it would be unwise to follow their preregistration in certain details, they may change it.
Allowing to deviate from the preregistration is controversial (XXX) because the authors might sift through all the possibilities and than select the most beneficial for their theory.
This argument, in its fullest consequence, might result in the same uncertainty we argued is appropriate for not preregistered studies.
However, if the authors offer a convincing argument for their deviation, their readers might be compelled to increase their uncertainty only a limited amount.
